---
title: "Coding the Matrix: Linear Algebra through Computer Science Applications"
author: "Zhiwei YANG (z3175089)"
date: "Friday, February 27, 2015"
output: html_document
---

## Language engines
### Use other languages in knitr
We can use any languages in __knitr__, including but not limited to `R`. So `R` and `Python` in the same report, brilliant!

```{r py_hello, engine='python'}
x = 'hello, python world'
print(x)
print(x.split(' '))
```

```{r package_options, include=FALSE}
# opts_knit$set(root.dir, 'D:/MyData/Personal/Dropbox/Coursera/Brown/matrix')
```

```{r sysinfo}
print(Sys.info())
print(Sys.info()['sysname'])
```


```{r global_options, include=FALSE}
# library(knitr)

if (Sys.info()['sysname'] == "Windows") {
  knitr::opts_chunk$set(engine='python') # Windows  
} else {
  knitr::opts_chunk$set(engine='python', engine.path='/usr/bin/python3') # Ubuntu (Linux)
}
```

```py
pandoc example
```

## Week 0: The Function and the Field
### The Function problems

#### Problem 1: tuple_sum(A, B)
```{r py_tuplesum}
from The_Function_problems import tuple_sum
A = [(1, 2), (10,20)]
B = [(3, 4), (30,40)]
print(tuple_sum(A, B))
```

#### Problem 2: inv_dict(d)
```{r py_invdict}
from The_Function_problems import inv_dict
print(inv_dict({'goodbye':  'au revoir', 'thank you': 'merci'}))
```

#### Problem 3: list of lists
```{r py_listoflists}
from The_Function_problems import row
print(row(10, 4))

comprehension_with_row = [row(i, 20) for i in range(15)]
print(comprehension_with_row)

comprehension_without_row = [[i+j for j in range(20)] for i in range(15)]
print(comprehension_without_row)
```

#### Functional Inverses
##### Ungraded problems
The first function is invertible because it is one-to-one and onto. The second function is not invertible. However, if we change the codomain $V$ to be the same as the domain $U$, it becomes invertible.

#### Functional Composition
##### Ungraded problems
The domain and codomain of $g(x)$ is both $\mathbb{R}^+$ so that $g\circ f$ is defined.

$f\circ g$ is defined and we have $(f\circ g)(1)=23,\,(f\circ g)(2)=22$ and $(f\circ g)(3)=21$.

#### Problem 4: probalities of a function
Note that the probability distribution is defined on $f(x)$'s __domain__, rather than codomain - $f(x)$ itself is deterministic.
$$\mathbb{P}(f(x)\in \{2, 4, 6\})=\mathbb{P}(x=1)+\mathbb{P}(x=3)+\mathbb{P}(x=5)=0.5+0.1+0.1=0.7$$
$$\mathbb{P}(f(x)\in \{3, 7\})=\mathbb{P}(x=2)+\mathbb{P}(x=6)=0.2+0.1=0.3$$

#### Problem 5: probalities of a function
Note that the probability distribution is defined on $g(x)$'s __domain__, rather than codomain - $g(x)$ itself is deterministic.
$$\mathbb{P}(g(x)=1)=\mathbb{P}(x\in\{1, 4, 7\})=\mathbb{P}(1)+\mathbb{P}(4)+\mathbb{P}(7)=0.2+0.1+0.1=0.4$$
$$\mathbb{P}(g(x)\in\{0,2\})=1-\mathbb{P}(g(x)=1)=1-0.4=0.6$$

### The Field problems
#### Problem 1: my_filter(L, num)
```{r py_myFilter}
from The_Field_problems import myFilter
print(myFilter([1,2,4,5,7],2))
print(myFilter([10,15,20,25],10))
```

#### Problem 2: my_lists(L)*
```{r py_my_lists}
from The_Field_problems import my_lists
print(my_lists([1,2,4]))
print(my_lists([0,3]))
```

#### Problem 3: my_function_composition(f,g)
```{r py_my_compo}
from The_Field_problems import myFunctionComposition

f = {0:'a',1:'b'}
g = {'a':'apple','b':'banana'}
print(myFunctionComposition(f,g))

a = {'x':24,'y':25}
b = {24:'twentyfour',25:'twentyfive'}
print(myFunctionComposition(a,b))
```

#### Problem 4: mySum(L)
```{r py_mySum}
from The_Field_problems import mySum
print(mySum([1,2,3,4]))
print(mySum([3,5,10]))
```

#### Problem 5: myProduct(L)
```{r py_myProduct}
from The_Field_problems import myProduct
print(myProduct([1,3,5]))
print(myProduct([-3,2,4]))
```

#### Problem 6: myMin(L)
```{r py_myMin}
from The_Field_problems import myMin
print(myMin([1,-100,2,3]))
print(myMin([0,3,5,-2,-5]))
```

#### Problem 7: myConcat(L)
```{r py_myConcat}
from The_Field_problems import myConcat
print(myConcat(['hello','world']))
print(myConcat(['what','is','up']))
print(myConcat([]))
```

#### Problem 8: myUnion(L)
```{r py_myUnion}
from The_Field_problems import myUnion
print(myUnion([{1,2},{2,3}]))
print(myUnion([set(),{3,5},{3,5}]))
```

##### Ungraded problems
1. `0`
2. `1`
3. `float("inf")` - infinity
4. `""` - empty string
5. `set()` - empty set

It is impossible to define the identity element for intersection operation in practice.

#### Problem 9: sum of complex numbers
```{r py_sumofcomplex}
print((3+1j)+(2+2j))
print((-1+2j)+(1-1j))
print((2+0j)+(-3+.001j))
print(4*(0+2j)+(.001+1j))
```

### Combining operations on complex numbers
#### Problem 10: transform(a, b, L)
```{r py_transform}
from The_Field_problems import transform
print(transform(3,2,[1,2,3]))
```

a. $2*(-i)*(z+(1+1i))=(-2i)z+(2-2i)$. Hence, $a=-2i$ and $b=2-2i$.
b. Scaling the real part and imaginary part by different factors is impossible!

#### Problem 11: arithmatic over GF(2)
```{r py_gf2}
from GF2 import one

print(one + one + one + 0)
print(one*one + 0*one + 0*0 + one*one)
print((one + one + one)*(one + one + one +one))
```

## Week 1: The Vector
### The Vector problems
#### Vector Addition Practice
#### Problem 1, 2, 3
Manually calculated.


### Expressing one $GF(2)$ vector as a sum of others
Manually calculated.

#### Problem 5
There are patterns in a, b,..., and f.

#### Problem 6
Dot product over GF(2). $x+1111$ holds because all three vectors have dot-product 0 with 1111 then add 0 to both sides of the equations.

```{r vecaddition}
from GF2 import one

```

#### Problem 7
Formulate equations using dot-product, i.e. come up with three vectors represented as lists.

#### Problem 8
For each of the following pairs of vectors over $\mathbb{R}$, evaluate the expression $\mathbf{u}\cdot\mathbf{v}$.

### Vector Class Homework
#### Problem 1: implement operations in class `Vec`
To invoke `cmd` terminal from a specific folder in Windows, either `Shift + Right Click` then select from menu or `Alt + D` to move to address bar then type `cmd`. From a console, make sure your current working directory is the one containing `vec.py`, and type
`python -m doctest vec.py`

Inline Python doctest results are as follows
```{r py_vec}
import doctest
doctest.testfile("vec.py")
```
